Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-12 text/manuscript-full-KS.qmd
full version

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. By showing which students are included versys excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

Having defined geography, the Market Segment Model sought to predict how student demand for a particular college varies by Geomarket, based on the characteristics of households in the Geomarket:

> For our analysis, we sought not a complex mathematical model, but a straightforward classification system that would track the pattern of SAT-score submissions to create a map of student choice. The Market Segment model that we developed was nothing more than a set of simple rules for disaggregating high school seniors into similar groups. The model worked because students, once so disaggregated, appeared to behave in remarkably consistent ways [@RN4982, p. 4].

SAT test-takers were categorized into one of four different *market segments* -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. <!-- 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a colleges which Geomarkets possess attractrive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

The thesis of Chapter 3 is that student demand for higher education is a function of social origin. For the authors (p. 33) state that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and that "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982 \[p. 33\] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time."\[\^7\] @RN4982 [p. 42] conclude that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations." The authors also find that these four variables predict student score-sending behavior at the Geomarket-level. However, Geomarkets differ in the relative abundance of students with particular socioeconomic characteristics, which has practical implications for recruiting. Therefore, @RN4982 [p. 44] recommend that colleges target Geomarkets with desirable compositions of socioeconomic characteristics in order to reach students from desired student market segments:

@@

The discussion of competition by @RN4982 exemplifies the concerns about correlation and homophily described by @RN4975. A correlational analysis of 1980 SAT score-sending patterns finds competition between colleges is defined by socioeconomic homophily. This homophily is presented as a naturally occurring phenomenon. Given these findings, @RN4982 recommend that colleges should target Geomarkets that contain a critical mass of students interested in peer-coleges, information that can be discerned from the Institutional Profiles (Appendix @tbl-sample-institutional-profile). In itself, @RN4982's Market Segment Model is a social science depiction of student demand -- akin to the status attainment model -- that does not consider historic, structural inequalities that cause observed patterns. However, by inscribing the Market Segment Model and Geomarkets into the EPS software, College Board amplified structural inequalities that contributed to homophily observed in 1980 SAT score-sending beahivor. In this way, commodification of the Market Segment Model and Geomarkets "engineers homophily" [@RN4975].

>>>>

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. By showing which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

Having defined geography, the Market Segment Model sought to predict how student demand for a particular college varies by Geomarket, based on the characteristics of households. The model created a straightforward but consistent classification system that grouped SAT test-takers into one of four different *market segments* -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc.

<!--# For our analysis, we sought not a complex mathematical model, but a straightforward classification system that would track the pattern of SAT-score submissions to create a map of student choice. The Market Segment model that we developed was nothing more than a set of simple rules for disaggregating high school seniors into similar groups. The model worked because students, once so disaggregated, appeared to behave in remarkably consistent ways [@RN4982, p. 4]. -->

<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

The thesis of Chapter 3 is that student demand for higher education is a function of social origin. For the authors (p. 33) state that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and that "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982 \[p. 33\] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time."\[\^7\] @RN4982 [p. 42] conclude that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations."

The authors also find that these four variables predict student score-sending behavior at the Geomarket-level. However, Geomarkets differ in the relative abundance of students with particular socioeconomic characteristics, which has practical implications for recruiting. Therefore, @RN4982 [p. 44] recommend that colleges target Geomarkets with desirable compositions of socioeconomic characteristics in order to reach students from desired student market segments:

@@

The discussion of competition by @RN4982 exemplifies the concerns about correlation and homophily described by @RN4975. A correlational analysis of 1980 SAT score-sending patterns finds competition between colleges is defined by socioeconomic homophily. This homophily is presented as a naturally occurring phenomenon. Given these findings, @RN4982 recommend that colleges should target Geomarkets that contain a critical mass of students interested in peer-colleges, information that can be discerned from the Institutional Profiles (Appendix @tbl-sample-institutional-profile). In itself, @RN4982's Market Segment Model is a social science depiction of student demand -- akin to the status attainment model -- that does not consider historic, structural inequalities that cause observed patterns. However, by inscribing the Market Segment Model and Geomarkets into the EPS software, College Board amplified structural inequalities that contributed to homophily observed in 1980 SAT score-sending beahivor. In this way, commodification of the Market Segment Model and Geomarkets "engineers homophily" [@RN4975].
